### 2020-03-13 thesis/content/chapter-graph-lta.tex
started with svm lta formulation proof

Based on this intuition we now give a generalized definition of \ac{lta} which applies to unstructured as well as structured input data.
We assume that all compositions are represented by graphs $G \in \mathcal{G}$;
an unstructured input is represented by a graph with one vertex per constituent ($\mathcal{V}_G = \{ v_{c_1}, \dots, v_{c_n} \}$) and no edges ($\mathcal{E}_G = \emptyset$).
Each composition has some target score $y_G \in \mathcal{Y}$ which could be a discrete class or continuous value.
An \textit{\ac{lta} model} $h: \mathcal{G} \to \mathcal{Y}$ assigns predictions $\hat{y}_G \in \mathcal{Y}$ to compositions $G$ which ideally correspond to the true score $y_G$.
Such a model must satisfy three criteria:
\begin{enumerate}[label=\textbf{\arabic*.}]
	\item \textbf{Decomposition:}
		A given composition $G$ must be decomposed into a set of constituents $c_{G,i}$ via a \textit{decomposition function} $\psi: \mathcal{G} \to \mathcal{P}(\mathcal{G})$.
		\begin{defn}
			$\psi$ is a \textit{decomposition function} iff.\ it splits a graph into a subset of its subgraphs, i.e.\ $\forall G \in \mathcal{G}: \forall c_{G,i} \in \psi(G): \exists s \in \mathcal{V}_G^{*}: c_{G,i} \equiv G[s]$.
		\end{defn}

@@

		Note that there are no explicit constituent weights in the existing unstructured \ac{lta} approaches (i.e.\ implicitly all $w_{G, i} = 1$) because the explicitly given constituents are assumed to be equally indicative of $y_G$.
		For structured data however, where the decomposition $\psi(G)$ is not given as part of the input, this assumption does not necessarily hold.
		By weighting the constituents, an \ac{lta} model can reduce the relevance or even ignore constituents that turn out to be irrelevant for the compositions target score $y_G$.
	\item \textbf{Aggregation:}
		Lastly a \textit{weighted aggregation function} $\mathcal{A}: {(\mathcal{Y} \times \mathbb{R})}^{*} \to \mathcal{Y}$ must be applied.
		It combines the multiset of local constituent predictions and weights into a single global composition prediction.
		\begin{defn}\label{defn:ltag:weighted-agg}
			We call $\mathcal{A}$ a \textit{weighted aggregation function} iff.\ it satisfies
			\begin{align*}
				\text{idempotency: } & \forall y \in \mathcal{Y}, w \in \mathbb{R}_{> 0}^n: \mathcal{A}({\ldblbrace (y, w_i) \rdblbrace}_{i=1}^n) = y \\
				\land\text{ zero invariance: } & \forall y_0 \in \mathcal{Y}, S = {\ldblbrace (y_i, w_i) \rdblbrace}_{i=1}^n: \mathcal{A}(S \cup \{ (y_0, 0) \}) = \mathcal{A}(S) \text{.}
			\end{align*}
		\end{defn}
		One such weighted aggregator is the \textit{weighted mean} $\wmean({\ldblbrace (y_i, w_i) \rdblbrace}_{i=1}^n) \coloneqq \sum_{i=1}^n w_i y_i$ which requires $\sum w_i = 1$ and $w_i \in [0, 1]$.
		Alternatively an unweighted aggregation function like $\min$, $\max$ or \ac{owa} also trivially satisfies \cref{defn:ltag:weighted-agg} if all inputs $y_i$ with $w_i = 0$ are filtered out and the weights for all remaining inputs are ignored.
\end{enumerate}
Based on the notion of decomposition, disaggregation and aggregation we can now define the concept of \textit{\ac{lta} formulations}.
\begin{defn}
	A model $h: \mathcal{G} \to \mathcal{Y}$ is in an \textit{\ac{lta} formulation} iff.\ it is expressed as
	\begin{align*}
		h(G) \coloneqq \mathcal{A}(\ldblbrace f(c_{G,i})\, |\, c_{G,i} \in \psi(G) \rdblbrace) \quad\text{with $\psi$, $f$ and $\mathcal{A}$ as defined above.}
	\end{align*}
\end{defn}
Note that every model $h: \mathcal{G} \to \mathcal{Y}$ has a trivial recursive \ac{lta} formulation by choosing $\psi(G) = \{ G \}$, $f(G) = (h(G), 1)$ and an arbitrary weighted aggregation function $\mathcal{A}$.
Those trivial \ac{lta} formulations do not split compositions into locally evaluated constituents and therefore intuitively do not fulfill the postulated localized explainability characteristic of \ac{lta}.
However since there is no commonly accepted formal criterion to decide whether a model's decisions are explainable~\cite{Lipton2018}, we do not attempt to strictly distinguish between \ac{lta} and non-\acs{lta} methods.
Instead the notion of \ac{lta} formulations should be seen as way to identify how ``\acs{lta}-like'' a model is:
\begin{itemize}
	\item \textbf{Negative extreme:}
		If a model $h$ only has trivial \ac{lta} formulations with the decomposition function $\psi(G) = \{ G \}$, it is not considered to be an \ac{lta} model.
	\item \textbf{Positive extreme:}
		If a model has an \ac{lta} formulation with a decomposition function that returns interpretable constituents, it is considered to be an \ac{lta} model.
		By definition this is true for the constituents $\psi(G) \coloneqq {\{ G[v_{c_i}] \}}_{v_{c_i} \in \mathcal{V}_G}$ of \ac{lta} methods for unstructured data.
	\item \textbf{In-between cases:}
		An \ac{lta} method for structured data must produce models that lie somewhere in-between the two extremes.
\end{itemize}

The more ``\acs{lta}-like'' a given model is, the stronger its bias towards locally explainable predictions, which in turn reduces the potential expressive power of the model.
\citet{Gilpin2018} describe this trade-off between explainability and expressive power in more detail.
When considering problem domains in which the true composition scores $y_G$ are accurately described by an \acs{lta}-like generative process, a less expressive \ac{lta}-like model could however generalize better than a more expressive non-\ac{lta} model.
This idea is captured by the so-called \textit{\ac{lta} assumption}.
\begin{defn}
	A problem domain $\mathcal{D}$ satisfies the \textit{\ac{lta} assumption} iff.\ there is an \ac{lta} method which produces models with an equal or lower out-of-sample error than the models produced by non-\acs{lta} methods for most training samples from $\mathcal{D}$.
\end{defn}
Due to the fuzziness of the class of \ac{lta} methods, the \ac{lta} assumption is naturally also a fuzzy concept.
Nonetheless evidence for its truthiness in a given domain $\mathcal{D}$ can be empirically obtained by comparing candidate \ac{lta} methods with the best known non-\ac{lta} method for $\mathcal{D}$, assuming that some cut-off condition for the required ``\ac{lta}-likeness'' of an \ac{lta} method is agreed upon.

@@

In \cref{sec:related:gcr:embed,sec:related:gcr:kernel} three different ways to map a given graph $G$ to a vector representation $\varphi(G) \in \mathbb{R}^d$ were described:
\begin{enumerate*}
	\item Fingerprint embeddings,
	\item skip-gram inspired embeddings and
	\item kernel embeddings
\end{enumerate*}.

>>>>

Based on this intuition we now give a generalized definition of \ac{lta} which applies to unstructured as well as structured input data.
We assume that all compositions are represented by graphs $G \in \mathcal{G}$;
an unstructured input is represented by a graph with one vertex per constituent ($\mathcal{V}_G = \{ v_{c_1}, \dots, v_{c_n} \}$) and no edges ($\mathcal{E}_G = \emptyset$).
Each composition has some target score $y_G \in \mathcal{Y}$ which could be a discrete class or continuous value.
An \textit{\ac{lta} model} $h: \mathcal{G} \to \mathcal{Y}$ assigns predictions $\hat{y}_G \in \mathcal{Y}$ to compositions $G$ which ideally correspond to the true score $y_G$.
Such a model must satisfy three criteria:
\begin{enumerate}[label=\textbf{\arabic*.}]
	\item \textbf{Decomposition:}
		A given composition $G$ must be decomposed into a set of constituents $c_{G,i}$ via a \textit{decomposition function} $\psi: \mathcal{G} \to \mathcal{P}(\mathcal{G})$.
		\begin{defn}\label{defn:ltag:decomp}
			$\psi$ is a \textit{decomposition function} iff.\ it splits a graph into a subset of its subgraphs, i.e.\ $\forall G \in \mathcal{G}: \forall c_{G,i} \in \psi(G): \exists s \in \mathcal{P}(\mathcal{V}_G): c_{G,i} = G[s]$.
		\end{defn}
		Note that the strict equality $c_{G,i} = G[s]$ is used in \cref{defn:ltag:decomp} instead of just requiring subgraph isomorphism ($c_{G,i} \simeq G[s]$) because a constituent $c_{G,i}$ represents a specific localized subgraph of a structured composition.

@@

		Note that there are no explicit constituent weights in the existing unstructured \ac{lta} approaches (i.e.\ implicitly all $w_{G, i} = 1$) because the explicitly given constituents are assumed to be equally indicative of $y_G$.
		For structured data however, where the decomposition $\psi(G)$ is not given as part of the input, this assumption does not necessarily hold.
		By weighting the constituents, an \ac{lta} model can reduce the relevance or even ignore constituents that turn out to be irrelevant for the compositions target score $y_G$.
	\item \textbf{Aggregation:}
		Lastly a \textit{weighted aggregation function} $\mathcal{A}: {(\mathcal{Y} \times \mathbb{R}_{\geq 0})}^{*} \to \mathcal{Y}$ must be applied.
		It combines the multiset of local constituent predictions and non-negative weights into a single global composition prediction.
		\begin{defn}\label{defn:ltag:weighted-agg}
			We call $\mathcal{A}$ a \textit{weighted aggregation function} iff.\ it satisfies
			\begin{align*}
				\text{idempotency: } & \exists c \geq 0: \forall y \in \mathcal{Y}, w \in \mathbb{R}_{\geq 0}^n\text{ s.t. }{\max w} \geq c: \mathcal{A}({\ldblbrace (y, w_i) \rdblbrace}_{i=1}^n) = y \\
				\land\text{ zero invar.: } & \forall y_0 \in \mathcal{Y}, S = {\ldblbrace (y_i, w_i) \rdblbrace}_{i=1}^n: \mathcal{A}(S \cup \{ (y_0, 0) \}) = \mathcal{A}(S) \text{.}
			\end{align*}
		\end{defn}
		{\setlength{\parskip}{0pt}The idempotency constraint requires aggregation functions to agree with uniform input scores $y$ if at least one input is weighted above some threshold $c$.
		The zero invariance constraint requires aggregation functions to ignore inputs with zero weight.}

		One example for such a weighted aggregator is the \textit{weighted mean} function $\wmean({\ldblbrace (y_i, w_i) \rdblbrace}_{i=1}^n) \coloneqq \sum_{i=1}^n w_i y_i$ which requires $\sum w_i = 1$ and $w_i \in [0, 1]$.
		Alternatively an unweighted aggregation function like $\min$, $\max$ or \ac{owa}\footnote{
			Even though \ac{owa} uses weights, it does not get those weights as part of the input and is therefore considered to be unweighted here.
		} also trivially satisfies \cref{defn:ltag:weighted-agg} if all inputs $y_i$ with $w_i = 0$ are filtered out and the weights for all remaining inputs are ignored.
\end{enumerate}
Based on the notion of decomposition, disaggregation and aggregation we can now define the concept of \textit{\ac{lta} formulations}.
\begin{defn}
	A model $h: \mathcal{G} \to \mathcal{Y}$ is in an \textit{\ac{lta} formulation} iff.\ it is expressed as
	\begin{align*}
		h(G) \coloneqq \mathcal{A}(\ldblbrace f(c_{G,i})\, |\, c_{G,i} \in \psi(G) \rdblbrace) \quad\text{with $\psi$, $f$ and $\mathcal{A}$ as defined above.}
	\end{align*}
\end{defn}
Note that every model $h: \mathcal{G} \to \mathcal{Y}$ has a trivial recursive \ac{lta} formulation by choosing $\psi(G) = \{ G \}$, $f(G) = (h(G), 1)$ and an arbitrary weighted aggregation function $\mathcal{A}$.
Those trivial \ac{lta} formulations do not split compositions into locally evaluated constituents and therefore intuitively do not fulfill the postulated localized explainability characteristic of \ac{lta}.
However since there is no commonly accepted formal criterion to decide whether a model's decisions are explainable~\cite{Lipton2018}, we do not attempt to strictly distinguish between \ac{lta} and non-\acs{lta} methods.
Instead the notion of \ac{lta} formulations should be seen as way to identify how ``\acs{lta}-like'' a model is:
\begin{itemize}
	\item \textbf{Negative extreme:}
		If a model $h$ only has trivial \ac{lta} formulations with the decomposition function $\psi(G) = \{ G \}$, it is not considered to be an \ac{lta} model.
	\item \textbf{Positive extreme:}
		If a model has an \ac{lta} formulation with a decomposition function that returns interpretable constituents, it is considered to be an \ac{lta} model.
		By definition this is true for the constituents $\psi(G) \coloneqq {\{ G[v_{c_i}] \}}_{v_{c_i} \in \mathcal{V}_G}$ of \ac{lta} methods for unstructured data.
	\item \textbf{In-between cases:}
		An \ac{lta} method for structured data must produce models that lie somewhere in-between the two extremes.
\end{itemize}

The more ``\acs{lta}-like'' a given model is, the stronger its bias towards locally explainable predictions, which in turn reduces the potential expressive power of the model.
\citet{Gilpin2018} describe this trade-off between explainability and expressive power in more detail.
When considering problem domains in which the true composition scores $y_G$ are accurately described by an \acs{lta}-like generative process, a less expressive \ac{lta}-like model could however generalize better than a more expressive non-\ac{lta} model.
This idea is captured by the so-called \textit{\ac{lta} assumption}.
\begin{defn}
	A problem domain $\mathcal{D}$ satisfies the \textit{\ac{lta} assumption} iff.\ there is an \ac{lta} method which produces models with an equal or lower out-of-sample error than the models produced by non-\acs{lta} methods for most training samples from $\mathcal{D}$.
\end{defn}
Due to the fuzziness of the class of \ac{lta} methods, the \ac{lta} assumption is naturally also a fuzzy concept.
Nonetheless evidence for its truthiness in a given domain $\mathcal{D}$ can be empirically obtained by comparing candidate \ac{lta} methods with the best known non-\ac{lta} method for $\mathcal{D}$, assuming that some cut-off condition for the required ``\ac{lta}-ness'' of an \ac{lta} method is agreed upon.

@@

In \cref{sec:related:gcr:embed,sec:related:gcr:kernel} three different ways to map a given graph $G$ to a vector $\varphi(G) \in \mathbb{R}^d$ were described:
\begin{enumerate*}
	\item Fingerprint embeddings,
	\item skip-gram inspired embeddings and
	\item kernel embeddings
\end{enumerate*}.
One common way to solve the \ac{gcr} problem via those embedding vectors is to train an \ac{svm} on them.
We will now see that \acp{svm} can be interpreted as \ac{lta} models if they are trained on so-called \textit{substructure component embeddings}.
\begin{defn}\label{defn:ltag:substruct-embedding}
	A graph embedding $\varphi: \mathcal{G} \to \mathbb{N}^{d}$ is called a \textit{substructure component embedding} iff.\ %
	$\forall i \in [d]: \exists S_i \in \mathcal{G}_{\mathit{ord}}, \gamma_i \in \mathbb{R}: \forall G \in \mathcal{G}: {{\varphi(G)}[i] = \mathit{contains}_{S_i}(G)} \lor {{\varphi(G)}[i] = \gamma_i \mathit{count}_{S_i}(G)}$,
	with $\mathit{contains}$ and $\mathit{count}$ as in \cfullref[ on ]{defn:related:wl-count-detect}.
\end{defn}
Intuitively \cref{defn:ltag:substruct-embedding} states that every embedding dimension must represent the count or presence of some ordered subgraph $S_i$ in the embedded graph $G$.
Based on this requirement we now proof the main theorem which shows the relation between \acp{svm} and \ac{lta}.
\begin{thm}
	A binary \ac{svm} classifier that takes substructure component embeddings $\varphi(G)$ as its input has an \ac{lta} formulation with a nontrivial decomposition function $\psi$.
\end{thm}
\begin{proof}
	Let $h: \mathcal{G} \to {\{ -1, +1 \}}$ be a binary graph classifier expressed as $h = h_{\text{\acs*{svm}}} \circ \varphi$ where $\varphi: \mathcal{G} \to \mathbb{N}_0^{d}$ is a substructure component embedding and $h_{\text{\acs*{svm}}}: \mathbb{R}^{d} \to {\{ -1, +1 \}}$ an \ac{svm} classifier.
	If $h$ is trained on a dataset $\mathcal{D}_{\text{train}} = {\{ (G_1, y_1), \dots, (G_N, y_N) \}}$, via the kernel trick it can be expressed as
	{\allowdisplaybreaks\begin{align*}
		h(G) &= \sgn\left( \sum_{j = 1}^{N} \alpha_j y_j {\langle \varphi(G), \varphi(G_j) \rangle} + b \right)
		\quad\text{for some $\alpha \in \mathbb{R}^N$ and $b \in \mathbb{R}$} \\
		&= \sgn\left( {\sum_{i=1}^{d} {\varphi(G)}[i] \underbrace{\left( \sum_{j = 1}^{N} \alpha_j y_j {\varphi(G_j)}[i] \right)}_{{} \eqqcolon z_i}} + b \right)
		\text{.}
	\end{align*}}

	By \cref{defn:ltag:substruct-embedding} each $\varphi(G)[i] > 0$ counts or indicates the presence of some ordered subgraph $S_i$ in $G$;
	more precisely:
	\begin{itemize}
		\item If $\varphi(G)[i] = \gamma_i \mathit{count}_{S_i}(G) > 0$, let $C_i \coloneqq {\{ G[s]\, |\, s \in \mathcal{P}(\mathcal{V}_G) \land G[s] \simeq S_i \}}$ be the $\mathit{count}_{S_i}(G)$ subgraphs counted by the $i$-th dimension.
		\item If $\varphi(G)[i] = \mathit{contains}_{S_i}(G) = 1$, let $C_i \coloneqq {\{  \}}$.
	\end{itemize}
	A substructure component embedding vector can therefore be interpreted as an encoding of the graph decomposition \[
		\psi(G) \coloneqq \bigcup_{i=1}^{d}
		\text{.}
	\]
	\begin{align*}
		h(G) &= \sgn\left( {\sum_{i=1}^{d} {\varphi(G)}[i] \underbrace{|z_i|}_{{} \eqqcolon \nu_i} \underbrace{\sgn(z_i)}_{{} \eqqcolon y_i}} + b \right)
	\end{align*}
\end{proof}
